Demographic history and patterns of selection in maize during and since domestication

### 2015-07-31 manuscript.tex
many updates

To ensure that this analysis is resilient to the intricacies of our sample, we investigated diversity surrounding the maize gene tga1, one of the few known instances of a hard sweep corresponding to a causal nonsynonymous mutuation in maize \cite{wang2015}. In the case of this known sweep, we observed a reduction in diversity surrounding the causitive substitution to levels much lower than seen for synonymous substitutions \textcolor{red}{(SUPP FIG)}, confirming that this method should be informative if hard sweeps were the prevaling mode of selection in maize. Still, to ensure that the approach is robust to potential shortcomings such as limited power in the face of background selection \cite{enard2014}, we sought to determine if patterns of diversity surrounding synonymous and non-synonymous substitutions differ when we restrict our analysis to the most- or least-conserved genes in maize. When we re-analyzed subsets of genes with the 10\% highest and 10\% lowest genomic evolutionary rate profile (GERP) score, putatively corresponding to genes undergoing the strongest and weakest purifying selection \cite{davydov2010, cooper2005, rodgers2015}, we again saw no significant difference in diversity surrounding synonymous and non-synonymous sites \textcolor{red}{(SUPP FIG)}.

@@

\begin{figure*}[t]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent
  past. Bootstrap-based 95\% confidence intervals are depicted via shading. Inset plots depict a smaller range on the x-axis. \label{fig:purify} \jri{should also add the singleton vs pi comparison within species as a supplemental figure. if $\pi$/singleton curves {\bf don't} match, will need to explain this suggests teosinte not in equilibrium.  good justification i think for msmc model, which we then use to say ``gee whiz, not perfect equilibrium but who cares because the big interesting demography is in maize.'' }}
\end{figure*}

\subsection{Demography of maize domestication}
To explore whether differences in purifying selection between maize
and teosinte can be explained by demographic processes, we
first estimated the parameters of maize domestication using large-scale
sequencing data involving 23 inbred maize
landraces and 13 teosinte inbred lines included in the HapMap 2 panel
\cite{chia2012} \textcolor{red}{(supplemental table here)}. The maize
lines were collected from across the Americas and the teosinte lines
came from central Mexico. Before estimating demography, we compared
the site frequency spectrum (SFS) in genic and non-genic regions, and
observed substantial differences in the evolution of these classes of
sites. For both maize and teosinte, the SFS within genes showed a
dearth of low-frequency alleles and Tajima's D
\cite{tajima1989} was therefore shifted
to more positive values \ref{fig:sffs}. This is consistent with the aforementioned purifying
selection and indicates that genic regions are not evolving
neutrally. This observation is consistent with previous research \cite{hufford2012}. Therefore, for demographic modeling we restricted analysis
to sites at least 5kb from gene start/stop postions.

@@

We utilized a complementary dataset of 4,021 maize landrace individuals collected
from across the Americas \textcolor{red}{(SeeDs reference here)} to
obtain a less downward-biased estimate of the current maize effective
population size. According to a singleton-based estimate of $\theta$
\cite{fu1993}, which was utilized since rapid post-bottleneck expansion
implies that rare alleles will most accurately reflect current maize
demography, we obtained the estimate \textcolor{red}{$N_e =
  992,713.5$}. Although less biased than the previously mentioned
estimate, we note that ascertainment bias of the
genotyping platform used to generate these data again biases this
estimate downward.

@@

Similarly, to complement our investigation of pairwise diversity surrounding substitutions, we studied patterns of singleton diversity surrounding synonymous and nonsynonymous substitutions in maize \textcolor{red}{(Supplemental Fig)}. A nearly identical pattern to that shown in Figure \ref{fig:hardSweeps} was observed. This further demonstrates that even the most recent selection patterns in maize are not dominated by hard sweeps.

\subsection{Purifying seleciton results are robust to soft-sweeps}
As a category, instances of soft-sweeps, or cases of selection on alleles that have already reached intermediate frequency before becoming advantageous, are difficult to identify \textcolor{red}{(refs here)}. However, this form of selection is likely to be frequent during crop domestication \textcolor{red}{(refs here)}. We were concerned that the disparate patterns of pairwise and singleton diversity in maize and teosinte, which seem to reflect differences in the efficiency of purifying selection due to demographic history, may instead reflect differences in the location and abundance of soft-sweeps. Therefore, we re-analyzed patterns of pairwise and singleton diversity in maize and teosinte including only a subset of sites that are least likely to correspond to soft-sweeps. To achieve this, we performed a genome-wide scan for soft or hard sweeps in maize or teosinte based on the H12 statistic \cite{garud2015}. H12 is sensitive to both hard and soft sweeps. Sites nearest to genes that displayed among the top 20\% of H12 values in either maize or teosinte were removed from our dataset, and both pairwise and singleton diversity at remaining sites, as a function of the distance to the nearest gene, was calculated \textcolor{red}{(Supp Fig)}. This subset of the data in no longer subject to the potentially confounding effects of soft-sweeps, and nearly identical patterns of pairwise and singleton diversity surrounding genes are observed. This provides further evidence that the patterns of diversity we observe in maize as compared to teosinte are the result of rapid population expansion as opposed to differing selective patterns in the domesticated and wild taxa.

@@

\jri{either we need to provide all scripts in a single github repo with a readme, or provide links to individual scripts in the methods as we go along}

\subsection{Plant materials}
Accessions studied were selected from the Maize HapMap2
panel \cite{chia2012, lemmon2014}. Principal component analysis was employed to ensure that closely related individuals were not included due to their potential to bias results \textcolor{red}{(maybe a supplemental figure here)}. \jri{yes supp. figure good idea}
Ultimately, 23 maize inbreds derived from a diverse assortment of landraces were selected for inclusion.
Thirteen teosinte inbred lines, all members of  subspecies Z. \emph{mays}
ssp. \emph{parviglumis}, were utilized.
Sequences were mapped to the maize B73 version 3 reference genome \cite{schnable2009} \url(ftp://ftp.ensemblgenomes.org/pub/plants/release-22/fasta/zea\_mays/dna/).

\subsection{Interpolating genetic position}
For many of the following analyses, physical position along a chromosome was a less relevant measure of map location than was genetic position.
Therefore, physical positions were converted to genetic positions by interpolating from the NAM genetic map \citex \jri{Wallace 2014 no?}, which provides a 1 cM resolution for physical to genetic conversion.
Within R \cite{R2014}, physical positions with corresponding genetic positions in the NAM map were used as anchors.
Physical positions in our dataset wihout corresponding genetic position were assigned genetic positions by scaling the anchored genetic positions according to the physical distance between the unlabeled position and the flanking anchors.

@@

\subsection{Demographic inference}
We used dadi \cite{gutenkunst2009} to estimate the parameters of maize
domestication based on diffusion approximation of the 2-dimensional
SFS for maize and teosinte. Our first step was to specify a model, as
shown in (Figure \ref{fig:bottleneck}). In words, our model began with an ancestral teosinte
population with effective population size $N_a$ individuals. The
ancestral population split into
two distinct populations, one of maize and the other of teosinte,
$T_b$ generations in the past. The maize population immediately shrank
to size $N_b$, and then grew exponentially to size $N_{maize}$
today. From the split until today, the teosinte effective population
size remained constant. Each year, $M_{mt}$ individuals migrated from
maize to teosinte, and $M_{tm}$ individuals migrated from teosinte to
maize.The free parameters of our model were $N_a, N_b, N_{maize}, T_b,
M_{mt}, \text{and } M_{tm} $. Python code specifying this model is
provided in \textcolor{red}{(SUPPLEMENT)}. We implimented 1,000 dadi
optimizations of this model using the ``dadi.Inference.optimize\_log\_fmin''
optimizer. In dadi optimizations,
initial values for $N_b, N_{maize}, T_b,
M_{mt}, \text{and } M_{tm}$, respectively, were randomly perturbed up
to a factor of two from 0.02, 3, 0.04, 1e-5, and 1e-5. Lower
bounds were set at 1e-7, 1e-7,  0, 1e-10, and
1e-10, respecively, while upper bounds were respectively set at 2,
200, 1, .001, and .001. $N_a$ did not have an initial value, upper, or
lower bounds because it is estimated from equilibrium ancestral theta
($\theta = 4N_a\mu$), which dadi determines automatically regardless
of model. Population sizes were converted from dadi's
units using a mutation rate estimate of 3e-8
\cite{clark2005}. Among the 1,000 dadi runs, parameters
from that with the lowest log-likelihood were taken as estimates of
the parameters of the domestication process.

Recent population expansion, as is observed for maize, most influences
the abundance of rare alleles. However, our demographic parameter
estimation using dadi suffered from a relatively small sample size of
23 maize individuals. Therefore, our approach likely underestimated
current maize effective population size. To address this, we utilized
a complimentary dataset of 4,021 maize landrace individuals collected
from across the Americas \textcolor{red}{(SeeDs reference here)} to
generate an independent estimate of
$N_{maize}$. These individuals were genotyped using GBS
\cite{elshire2011}, a protocol that suffers from a high rate of
missing data \cite{beissinger2013}. To minimiaze biases imposed from
missing data, we used R to isolate only those SNPs that
were observed in at least 1,500 individuals and subsequently projected the
SFS down to a sample of 500 individuals. Then, we utilized the singleton-based estimate of
theta ($\theta = 4N_e\mu = \frac{S}{L}$) to estimate
modern maize $N_e$ \cite{fu1993}, where $S$ is the total number of
singletons and $L$ is the total length of DNA sequenced. Based on the
GBS protocol implemented \cite{elshire2011}, $L$ is equal to the total
number of SNPs in the dataset.

\subsection{Evaluating diversity around substitutions}
To investigate diversity around substitutions, maize and teosinte pairwise diversity was first
calculated in 1,000 kb non-overlapping windows using ANGSD
\cite{korneliussen2014}. This was performed separately for both maize and
teosinte, using the same filters as employed for estimating the
SFS. Next, SNPs and genotypes among maize, teosinte, and \emph{Tripsacum} were called. \emph{Tripsacum} bam files
were downloaded from \url{}(TRIPSICUM FILES), and then all SNPs with a
p-value less than 1e-6 were called using ANGSD. Quality filters were
as the same as before, and genotypes were only called when the
posterior probability was above 0.95. From the set of called SNPs and
genotypes, substitutions between maize and tripsicum, as well as
between teosinte and \emph{Tripsacum} were identified using R \cite{R2014} as all positions with
no more than 20\% missing data for which every maize or teosinte
allele differed from the observed \emph{Tripsacum} allele. At each class of
substitution, effects were estimated using the ensembl variant effects
predictor \cite{mclaren2010}.

For each diversity window with at least 100 bps observed, the distance from the window center to the
nearest synonymous and nonsynonymous (missense) substitution was
computed. Then, following the methods of \cite{hernandez2011}, a loess
curve was plotted for diversity values against the distance to the
nearest synonymous or nonsynonymous substitution. A span of 0.01 was
utilized. Unlike
\cite{sattath2011}, we did not fit separate loess curves in the up- and
down-stream directions, but instead fit single curves encompassing
both directions.

@@

@@

\begin{acknowledgments}
Various thankyous will be in order. \jri{Coop and his lab. funding from NSF Biology of Rare alleles -- need grant number.}
\end{acknowledgments}

>>>>

To ensure that this analysis is resilient to the intricacies of our sample, we investigated diversity surrounding the maize gene tga1, one of the few known instances of a hard sweep corresponding to a causal nonsynonymous mutuation in maize \cite{wang2015}. In the case of this known sweep, we observed a reduction in diversity surrounding the causitive substitution to levels much lower than seen for synonymous substitutions (Figure S1), confirming that this method should be informative if hard sweeps were the prevaling mode of selection in maize. Still, to ensure that the approach is robust to potential shortcomings such as limited power in the face of background selection \cite{enard2014}, we sought to determine if patterns of diversity surrounding synonymous and non-synonymous substitutions differ when we restrict our analysis to the most- or least-conserved genes in maize. When we re-analyzed subsets of genes with the 10\% highest and 10\% lowest genomic evolutionary rate profile (GERP) score, putatively corresponding to genes undergoing the strongest and weakest purifying selection \cite{davydov2010, cooper2005, rodgers2015}, we again saw no significant difference in diversity surrounding synonymous and non-synonymous sites (Figure S2).

@@

\begin{figure*}[t]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent
  past. Bootstrap-based 95\% confidence intervals are depicted via shading. Inset plots depict a smaller range on the x-axis. \label{fig:purify}
  %\jri{should also add the singleton vs pi comparison within species as a supplemental figure. if $\pi$/singleton curves {\bf don't} match, will need to explain this suggests teosinte not in equilibrium.  good justification i think for msmc model, which we then use to say ``gee whiz, not perfect equilibrium but who cares because the big interesting demography is in maize.''}
  }
\end{figure*}

\subsection{Demography of maize domestication}
To explore whether differences in purifying selection between maize
and teosinte can be explained by demographic processes, we
first estimated the parameters of maize domestication using large-scale
sequencing data involving 23 inbred maize
landraces and 13 teosinte inbred lines included in the HapMap 2 panel
\cite{chia2012} \textcolor{red}{(supplemental table here)}. The maize
lines were collected from across the Americas and the teosinte lines
came from central Mexico. Before estimating demography, we compared
the site frequency spectrum (SFS) in genic and non-genic regions, and
observed substantial differences in the evolution of these classes of
sites. For both maize and teosinte, the SFS within genes showed a
dearth of low-frequency alleles and Tajima's D
\cite{tajima1989} was therefore shifted
to more positive values \ref{fig:tajima1989}. This is consistent with the aforementioned purifying
selection and indicates that genic regions are not evolving
neutrally. This observation is consistent with previous research \cite{hufford2012}. Therefore, for demographic modeling we restricted analysis
to sites at least 5kb from gene start/stop postions.

@@

We utilized a complementary dataset of 4,021 maize landrace individuals collected
from across the Americas \textcolor{red}{(SeeDs reference here)} to
obtain a less downward-biased estimate of the current maize effective
population size. According to a singleton-based estimate of $\theta$
\cite{fu1993}, which was utilized since rapid post-bottleneck expansion
implies that rare alleles will most accurately reflect current maize
demography, we obtained the estimate $N_e =
  992,713.5$. Although less biased than the previously mentioned
estimate, we note that ascertainment bias of the
genotyping platform used to generate these data again biases this
estimate downward.

@@

Similarly, to complement our investigation of pairwise diversity surrounding substitutions, we studied patterns of singleton diversity surrounding synonymous and nonsynonymous substitutions in maize (Figure S3). A nearly identical pattern to that shown in Figure \ref{fig:hardSweeps} was observed. This further demonstrates that even the most recent selection patterns in maize are not dominated by hard sweeps.

\subsection{Purifying seleciton results are robust to soft-sweeps}
As a category, instances of soft-sweeps, or cases of selection on alleles that have already reached intermediate frequency before becoming advantageous, are difficult to identify \textcolor{red}{(refs here)}. However, this form of selection is likely to be frequent during crop domestication \textcolor{red}{(refs here)}. We were concerned that the disparate patterns of pairwise and singleton diversity in maize and teosinte, which seem to reflect differences in the efficiency of purifying selection due to demographic history, may instead reflect differences in the location and abundance of soft-sweeps. Therefore, we re-analyzed patterns of pairwise and singleton diversity in maize and teosinte including only a subset of sites that are least likely to correspond to soft-sweeps. To achieve this, we performed a genome-wide scan for soft or hard sweeps in maize or teosinte based on the H12 statistic \cite{garud2015}. H12 is sensitive to both hard and soft sweeps. Sites nearest to genes that displayed among the top 20\% of H12 values in either maize or teosinte were removed from our dataset, and both pairwise and singleton diversity at remaining sites, as a function of the distance to the nearest gene, was calculated (Figure S4). This subset of the data in no longer subject to the potentially confounding effects of soft-sweeps, and nearly identical patterns of pairwise and singleton diversity surrounding genes are observed. This provides further evidence that the patterns of diversity we observe in maize as compared to teosinte are the result of rapid population expansion as opposed to differing selective patterns in the domesticated and wild taxa.

@@

  \subsection{BASH, R, and Python scripts}
All analysis scripts are available in an online repository at \textcolor{red}{REPO ADDRESS HERE.}

\subsection{Plant materials}
Accessions studied were selected from the Maize HapMap2
panel \cite{chia2012, lemmon2014}. Principal component analysis was employed to ensure that closely related individuals were not included due to their potential to bias results \textcolor{red}{(Figure S-PCA)}. Ultimately, 23 maize inbreds derived from a diverse assortment of landraces were selected for inclusion.
Thirteen teosinte inbred lines, all members of  subspecies Z. \emph{mays}
ssp. \emph{parviglumis}, were utilized.
Sequences were mapped to the maize B73 version 3 reference genome \cite{schnable2009} (ftp://ftp.ensemblgenomes.org/pub/plants/release-22/fasta/zea\_mays/dna/) as described by \cite{chia2012}.

\subsection{Interpolating genetic position}
For most of our analyses, physical position along a chromosome was a less relevant measure of map location than genetic position.
Physical positions were converted to genetic positions by interpolating from the NAM genetic map \cite{glaubitz2014}, which provides a 1 cM resolution for physical to genetic conversion.
Within R \cite{R2014}, physical positions with corresponding genetic positions in the NAM map were used as anchors.
Physical positions in our dataset wihout corresponding genetic position were assigned genetic positions by scaling the anchored genetic positions according to the physical distance between the unlabeled position and the flanking anchors.

@@

\subsection{Demographic inference}
We used dadi \cite{gutenkunst2009} to estimate the parameters of maize
domestication based on diffusion approximation of the 2-dimensional
SFS for maize and teosinte. Our first step was to specify a model, as
shown in (Figure \ref{fig:bottleneck}). In words, our model began with an ancestral teosinte
population with effective population size $N_a$ individuals. The
ancestral population split into
two distinct populations, one of maize and the other of teosinte,
$T_b$ generations in the past. The maize population immediately shrank
to size $N_b$, and then grew exponentially to size $N_{maize}$
today. From the split until today, the teosinte effective population
size remained constant. Each year, $M_{mt}$ individuals migrated from
maize to teosinte, and $M_{tm}$ individuals migrated from teosinte to
maize.The free parameters of our model were $N_a, N_b, N_{maize}, T_b,
M_{mt}, \text{and } M_{tm} $. We implimented 1,000 dadi
optimizations of this model using the ``dadi.Inference.optimize\_log\_fmin''
optimizer. In dadi optimizations,
initial values for $N_b, N_{maize}, T_b,
M_{mt}, \text{and } M_{tm}$, respectively, were randomly perturbed up
to a factor of two from 0.02, 3, 0.04, 1e-5, and 1e-5. Lower
bounds were set at 1e-7, 1e-7,  0, 1e-10, and
1e-10, respecively, while upper bounds were respectively set at 2,
200, 1, .001, and .001. $N_a$ did not have an initial value, upper, or
lower bounds because it is estimated from equilibrium ancestral theta
($\theta = 4N_a\mu$), which dadi determines automatically regardless
of model. Population sizes were converted from dadi's
units using a mutation rate estimate of 3e-8
\cite{clark2005}. Among the 1,000 dadi runs, parameters
from the run with the lowest log-likelihood were taken as estimates of
the parameters of the domestication process.

Recent population expansion, as is observed for maize, most influences
the abundance of rare alleles. However, our demographic parameter
estimation using dadi suffered from a relatively small sample size of
23 maize individuals. Therefore, our approach likely underestimated
current maize effective population size. To address this, we utilized
a complimentary dataset of 4,021 maize landrace individuals collected
from across the Americas \textcolor{red}{(SeeDs reference here)} to
generate an independent estimate of
$N_{maize}$. These individuals were genotyped using GBS
\cite{elshire2011}, a protocol that suffers from a high rate of
missing data \cite{beissinger2013}. To minimize biases imposed from
missing data, we used R to isolate only those SNPs that
were observed in at least 1,500 individuals and subsequently projected the
SFS down to a sample of 500 individuals. Then, we utilized the singleton-based estimate of
theta ($\theta = 4N_e\mu = \frac{S}{L}$) to estimate
modern maize $N_e$ \cite{fu1993}, where $S$ is the total number of
singletons and $L$ is the total length of DNA sequenced. Based on the
GBS protocol implemented \cite{elshire2011}, $L$ is equal to the total
number of SNPs in the dataset.

\subsection{Evaluating diversity around substitutions}
To investigate diversity around substitutions, maize and teosinte pairwise diversity was first
calculated in 1,000 kb non-overlapping windows using ANGSD
\cite{korneliussen2014}. This was performed separately for both maize and
teosinte, using the same filters as employed for estimating the
SFS. Next, SNPs and genotypes among maize, teosinte, and \emph{Tripsacum} were called. \emph{Tripsacum} bam files
were downloaded from \textcolor{red}{(TRIPSICUM FILES)}, and then all SNPs with a
p-value less than 1e-6 were called using ANGSD. Quality filters were
as the same as before, and genotypes were only called when the
posterior probability was above 0.95. From the set of called SNPs and
genotypes, substitutions between maize and tripsicum, as well as
between teosinte and \emph{Tripsacum} were identified using R \cite{R2014} as all positions with
no more than 20\% missing data for which every maize or teosinte
allele differed from the observed \emph{Tripsacum} allele. At each
substitution, effects were estimated using the ensembl variant effects
predictor \cite{mclaren2010}.

For each diversity window with at least 100 bps observed, the distance from the window center to the
nearest synonymous and missense substitution was
computed. Then, following the methods of \cite{hernandez2011}, a loess
curve was plotted for diversity values against the distance to the
nearest synonymous or nonsynonymous substitution. A span of 0.01 was
utilized.

To verify that this approach is robust to the criticisms of \cite{enard2014}, which involve reduced power at highly conserved sites, we applied the method to subsets of sites that included only positions nearest to the  most- or least-conserved 10\% of genes data. This was achieved by assigning every gene a GERP score corresponding to the average score observed across it \cite{rodgers2015}. Next, genes ing the top and bottom 10\% quantiles of GERP score were isolated. Two datasets were created, each including all diversity windows who's nearest gene was in the high or low GERP score sets. Finally, Loess fits of diversity vs. distance to gene were computed from sites within each GERP score.

@@

Additionally, we performed a parallel implementation of the above analysis after excluding sites potentially experiencing positive selection in the form of either hard- or soft-sweeps. To isolate these sites, we performed genome-scans separately for maize and teosinte based on the H12 statistic \cite{garud2015}. SNPs within each taxa were called in ANGSD \cite{korneliussen2014} using the previously specified SNP calling parameters. H12 was computed with a window size of 200 SNPs and a jump of 25 SNPs. H12 windows in the highest 20\% for maize or teosinte were identified and considered as regions most likely to be under positive selection. Finally, levels of pairwise and singleton diversity vs. distance to gene was computed using only sites that were not nearest to selected sites in either maize or teosinte.

@@

\begin{acknowledgments}
We are indebted to Graham Coop and Simon Aeshbacher for their constructive input during this study. Funding was provided by \textcolor{red}{NSF Grant number ZZZZZZZZZ}.
\end{acknowledgments}
